Trauma treatment delay

### 2023-09-15 manuscriptlongversion.Rmd
intro expand

---
title: Trauma treatment delay
subtitle: A retrospective cohort study on the factors affecting the delays in trauma care
author: Anton Wasielewski
bibliography: bibliography.bib
csl: bmcemerg.csl
---

@@

Trauma systems are infrastructures that exist to provide and optimize care for injured patients starting with injury recognition and triage, transport to appropriate trauma center, inpatient care and outpatient follow-up. Beyond the clinical side, it works with outreach, education and advocacy, data collection through registries, research, funding, and disaster preparedness and response. A comprehensive and functioning trauma system requires strong leadership and engagement at the trauma center, regional and national level [@connolly_woo_lampron_perry_2018]. This system is crucial to provide care for trauma patients, both in reducing morbidity and mortality in this patient group. Earlier studies from sweden have shown that treating severely injured patients at one of these establishments is associated with a 41% lower adjusted 30-day mortality rate compared to being treated at a non trauma care center due to them being more capable of treating these patients, with patients with higher ISS scores having an even higher protective effect with \>90% for the most critically injured patients [@Candefjord_Asker_Caragounis_2020]. Other studies have shown similar results, showing that treating injured patients at trauma centers is associated with a 15-30% decrease in mortality [@Moran_Lecky_Bouamra_Lawrence_Edwards_Woodford_Willett_Coats_2018][@elso_Tepas_Langland-Orban_Pracht_Papa_Lottenberg_Flint_2006][@MacKenzie_Rivara_Jurkovich_Nathens_Frey_Egleston_Salkever_Scharfstein_2006][@Haas_Stukel_Gomez_Zagorski_De_Mestral_Sharma_Rubenfeld_Nathens_2012].

@@

Trauma patients brought to trauma centers may receive immediate or delayed treatment depending on what protocols are present at the facility and how they are being followed [@ryb2012delayed]. Among the preventable or possibly preventable deaths in trauma patients, a group determined by the Combined Trauma Death Review Committee, delay in treatment has been identified as a major error contributing to death, found in up to 52.9% of patients in said group according to studies [@Teixeira_Inaba_Hadjizacharia_Brown_Salim_Rhee_Browder_Noguchi_Demetriades_2007]. In the case of a delayed trauma team activation (DTTA) it can often be attributed to insensitivity in triage protocol and failure to recognize injury patterns [@Hoyt_Hollingsworth-Fridlund_Winchell_Simons_Holbrook_Fortiage_1994]. Studies have also shown that different factors such as age over 55, nonwhite ethnicity, blunt assault among others was linked to a higher risk of DTTA [@ryb2012delayed].

>>>>

---
title: "Trauma treatment delay"
author: "Anton Wasielewski"
subtitle: A retrospective cohort study on the factors affecting the delays in trauma
  care
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
csl: bmcemerg.csl
bibliography: bibliography.bib
---

@@

Trauma systems are infrastructures that exist to provide and optimize care for injured patients starting with injury recognition and triage, transport to appropriate trauma center, inpatient care and outpatient follow-up. Beyond the clinical side, it works with outreach, education and advocacy, data collection through registries, research, funding, and disaster preparedness and response. A comprehensive and functioning trauma system requires strong leadership and engagement at the trauma center, regional and national level [@connolly_woo_lampron_perry_2018]. This system is crucial to provide care for trauma patients, both in reducing morbidity and mortality in this patient group. Earlier studies from Sweden have shown that treating severely injured patients at one of these establishments is associated with a 41% lower adjusted 30-day mortality rate compared to being treated at a non trauma care center due to them being more capable of treating these patients, with patients with higher ISS scores having an even higher protective effect with \>90% for the most critically injured patients [@Candefjord_Asker_Caragounis_2020]. Other studies have shown similar results, showing that treating injured patients at trauma centers is associated with a 15-30% decrease in mortality [@Moran_Lecky_Bouamra_Lawrence_Edwards_Woodford_Willett_Coats_2018][@elso_Tepas_Langland-Orban_Pracht_Papa_Lottenberg_Flint_2006][@MacKenzie_Rivara_Jurkovich_Nathens_Frey_Egleston_Salkever_Scharfstein_2006][@Haas_Stukel_Gomez_Zagorski_De_Mestral_Sharma_Rubenfeld_Nathens_2012]. During the last three decades, the introduction of trauma systems has lowered the incidence of preventable death from approximately 25% to less than 5%. This is attributed to the improvements in care for acute brain injuries and bleeding control. The incidence of late death because of sepsis and multiple organ failure, possibly a result of better and earlier resuscitation [@Asensio_Trunkey_2008].

@@

Trauma patients brought to trauma centers may receive immediate or delayed treatment depending on what protocols are present at the facility and how they are being followed [@ryb2012delayed]. Among the preventable or possibly preventable deaths in trauma patients, a group determined by the Combined Trauma Death Review Committee, delay in treatment has been identified as a major error contributing to death, found in up to 52.9% of patients in said group [@Teixeira_Inaba_Hadjizacharia_Brown_Salim_Rhee_Browder_Noguchi_Demetriades_2007]. In the case of a DTTA it can often be attributed to insensitivity in triage protocol and failure to recognize injury patterns [@Hoyt_Hollingsworth-Fridlund_Winchell_Simons_Holbrook_Fortiage_1994]. Studies have also shown that different factors such as age over 55, non-white ethnicity, blunt assault among others was linked to a higher risk of DTTA [@ryb2012delayed].
